---
title: "R Notebook"
output: html_notebook
---

# Statistical analysis for PHS scotland data

## Winter vs summer each year
- number of admissions
- proportion of beds
- length of stay

Compare mean from both


```{r}
library(tidyverse)
library(infer)
```


```{r}
age_sex_clean <- read_csv("data/clean_data/age_sex_clean.csv")
```

```{r}
age_sex_clean %>% 
  group_by(quarter) %>% 
  summarise(mean_episodes = mean(episodes)) 
  
```

```{r}
a_e_clean <- read_csv("raw_data/clean_datawaiting_times_clean.csv")
```



```{r}
a_e_clean %>% 
  ggplot(aes(x = NumberOfAttendancesAggregate))+
  geom_histogram()
```

```{r}
a_e_clean %>% 
  filter(str_detect(Season, "Winter$"))
```

```{r}
a_e_clean %>% 
  filter(str_detect(Season, "Winter$")) %>% 
   ggplot(aes(x = NumberOfAttendancesAggregate))+
  geom_histogram()
```

```{r}
winter %>% 
  summarise(mean = mean(NumberOfAttendancesAggregate))
```


```{r}
a_e_clean_winter<- a_e_clean %>%
  mutate(winter = case_when(Season == "DWinter" ~ "Winter",
                            TRUE ~ "Rest of Year"))
```



```{r}
a_e_clean_winter %>% 
  group_by(winter) %>% 
  summarise(mean_admissions = mean(NumberOfAttendancesAggregate),
            mean_discharge_home = mean(DischargeDestinationResidence, na.rm = TRUE),
            mean_transfer = mean(DischargeDestinationTransfer, na.rm = TRUE))
```

```{r}
a_e_clean_winter %>% 
  group_by(winter) %>% 
  ggplot(aes(x = NumberOfAttendancesAggregate, y = winter))+
  geom_boxplot()+
  labs(
    title = "A&E attendances"
  )
```

```{r}
a_e_clean_winter %>% 
  group_by(winter) %>% 
  ggplot(aes(x = DischargeDestinationResidence, y = winter))+
  geom_boxplot()+
  labs(
    title = "Patients discharged home")
```

```{r}
a_e_clean_winter %>% 
  group_by(winter) %>% 
  ggplot(aes(x = DischargeDestinationTransfer, y = winter))+
  geom_boxplot()+
  labs(
    title = "Patients transfered")
```


```{r}
a_e_clean %>% 
  group_by(Season) %>% 
  summarise(mean_admissions = mean(NumberOfAttendancesAggregate))
```
# Test 1: Are the number of admissions in winter higher than in the rest of the year?

## Step 1: generate hypothesis and significance level 

H0: mean(winter admissions) = mean(rest of year admissions)
$$H_0:  \mu_{\textrm{winter admissions}} -  \mu_{\textrm{rest of year admissions}} <=0$$
H1: mean(winter admissions) > mean(summer admissions)
$$H_1:  \mu_{\textrm{winter admissions}} -  \mu_{\textrm{rest of year admissions}}\ > 0$$
significance level = 0.05

## Step 2: Data wrangling

```{r}
a_e_clean_winter<- a_e_clean %>%
  mutate(winter = case_when(Season == "DWinter" ~ "Winter",
                            TRUE ~ "Rest of Year"))
```

Quick look at means
```{r}
a_e_clean_winter %>% 
  group_by(winter) %>% 
  summarise(mean_admissions = mean(NumberOfAttendancesAggregate, na.rm = TRUE),
            mean_discharge_home = mean(DischargeDestinationResidence, na.rm = TRUE),
            mean_transfer = mean(DischargeDestinationTransfer, na.rm = TRUE))
```

```{r}
a_e_clean_winter %>% 
  filter(winter == "Winter") %>% 
  ggplot(aes(x = NumberOfAttendancesAggregate))+
  geom_histogram()
```


```{r}
a_e_clean_winter %>% 
  filter(winter != "Winter") %>% 
  ggplot(aes(x = NumberOfAttendancesAggregate))+
  geom_histogram()
```
```{r}
a_e_clean_winter %>% 
  count(winter)
```
## Step 3: Calculate observed statistic (mean for both groups, winter - rest of year)


```{r}
observed_stat <- a_e_clean_winter %>% 
  specify(response = NumberOfAttendancesAggregate, explanatory = winter) %>% 
  calculate(stat = "diff in means", order = c("Winter", "Rest of Year"))

observed_stat
```
## Step 4: generate null distribution and visualise
  
```{r}
null_distribution <- a_e_clean_winter %>% 
  specify(response = NumberOfAttendancesAggregate, explanatory = winter) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("Winter", "Rest of Year"))
```

```{r}
null_distribution %>% 
  visualise(bins = 25)+
  shade_p_value(obs_stat = observed_stat$stat, 
                direction = "greater")
```
## Step 5: Get p-value and interpret

```{r}
null_distribution %>% 
  get_p_value(obs_stat = observed_stat$stat, 
                direction = "greater")
```
p-value is much higher than our significance level and therefore we fail to reject the H0 in favour of the Ha. ie the number of admissions in winter is equal or less than the mean number of admissions for the rest of the year.  


# Test 2: Rate of discharged home
Is there a higher discharge home rate in the summer compared to winter?

## Step1: Generate hypothesis
H0: mean(winter discharge home) = mean(rest of year discharge home)
$$H_0:  \mu_{\textrm{rest of year discharge home}} -  \mu_{\textrm{winter discharge home}} <=0$$
H1: higher mean discharge home in rest of year compared to winter )
$$H_1:  \mu_{\textrm{rest of year discharged home}} -  \mu_{\textrm{winter discharged home}}>0$$
significance level = 0.05

## Step 2: Data wrangling
Done - use a_e_clean_winter
Quick look at boxplot
```{r}
a_e_clean_winter %>% 
  group_by(winter) %>% 
  ggplot(aes(x = DischargeDestinationResidence, y = winter))+
  geom_boxplot()+
  labs(
    title = "Patients discharged home")
```

## Step 3: Calculate observed statistic (mean for both groups, subtract Winter from rest of year)


```{r}
obs_stat <- a_e_clean_winter %>% 
  specify(response = DischargeDestinationResidence, explanatory = winter) %>% 
  calculate(stat = "diff in means", order = c("Rest of Year", "Winter"))

obs_stat
```
## Step 4: generate null distribution and visualise
  
```{r}
null_distribution <- a_e_clean_winter %>% 
  specify(response = DischargeDestinationResidence, explanatory = winter) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("Rest of Year", "Winter"))
```

```{r}
null_distribution %>% 
  visualise(bins = 25)+
  shade_p_value(obs_stat = obs_stat$stat, 
                direction = "greater")
```
## Step 5: Get p value and interpret
```{r}
null_distribution %>% 
  get_p_value(obs_stat = obs_stat$stat, 
                direction = "greater")
```

p-value is less than our significance level and therefore we reject the H0 in favour of the Ha: ie there is higher discharge to home rate in the summer compared to winter.


# Test 3: Is the winter rate of admissions different pre and post covid?

Are the number of admissions in winter higher than in the rest of the year before Covid?

H0: mean(winter admissions pre_covid) = mean(rest of year admissions pre covid)
$$H_0:  \mu_{\textrm{winter admissions pre-covid}} -  \mu_{\textrm{rest of year admissions pre-covid}} <=0$$
H1: mean(winter admissions pre-covid) > mean(summer admissions pre-covid)
$$H_1:  \mu_{\textrm{winter admissions pre-covid}} -  \mu_{\textrm{rest of year admissions pre-covid}}\ > 0$$
## Step 2: Data wrangling and quick explore
```{r}
ae_winter_covid <- a_e_clean_winter %>%
  mutate(covid = case_when(str_detect(Quarter, "^201") ~ "pre_covid",
                            TRUE ~ "post_covid"))
```

```{r}
ae_winter_precovid <- ae_winter_covid %>% 
  filter(covid == "pre_covid")
```



```{r}
ae_winter_precovid %>% 
  group_by(winter) %>% 
  summarise(mean_admissions = mean(NumberOfAttendancesAggregate),
            mean_discharge_home = mean(DischargeDestinationResidence, na.rm = TRUE),
            mean_transfer = mean(DischargeDestinationTransfer, na.rm = TRUE))
```

## Step 3: Calculate observed stat (mean admissions pre-covid)

```{r}
obs_stat_precovid <- ae_winter_precovid %>% 
  specify(response = NumberOfAttendancesAggregate, explanatory = winter) %>% 
  calculate(stat = "diff in means", order = c("Winter", "Rest of Year"))

obs_stat_precovid
```
## Step4:  Generate null distribution and visualise
```{r}
null_dist_precovid <- ae_winter_precovid %>% 
  specify(response = NumberOfAttendancesAggregate, explanatory = winter) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("Winter", "Rest of Year"))
```

```{r}
null_dist_precovid %>% 
  visualise(bins = 25)+
  shade_p_value(obs_stat = obs_stat_precovid$stat, 
                direction = "greater")
```
## Step 5: get p-value and interpret

```{r}
null_dist_precovid %>% 
  get_p_value(obs_stat = obs_stat_precovid$stat, 
                direction = "greater")
```

Fail to reject the null hypothesis, ie the number of admissions to A&E in winter is the same as, or less than, admissions in the rest of the year. 


# Test 4: post covid admission rates

Are the number of admissions in winter higher than in the rest of the year after Covid?

H0: mean(winter admissions post_covid) = mean(rest of year admissions post covid)
$$H_0:  \mu_{\textrm{winter admissions post-covid}} -  \mu_{\textrm{rest of year admissions post-covid}} <=0$$
H1: mean(winter admissions post-covid) > mean(summer admissions post-covid)
$$H_1:  \mu_{\textrm{winter admissions post-covid}} -  \mu_{\textrm{rest of year admissions post-covid}}\ > 0$$
## Step 2: Data wrangling and quick explore

```{r}
ae_winter_postcovid <- ae_winter_covid%>% 
  filter(covid == "post_covid")
```


```{r}
ae_winter_postcovid %>% 
  group_by(winter) %>% 
  summarise(mean_admissions = mean(NumberOfAttendancesAggregate),
            mean_discharge_home = mean(DischargeDestinationResidence, na.rm = TRUE),
            mean_transfer = mean(DischargeDestinationTransfer, na.rm = TRUE))
```
##Step 3: Calculate observed stat

```{r}
obs_stat_postcovid <- ae_winter_postcovid%>% 
  specify(response = NumberOfAttendancesAggregate, explanatory = winter) %>% 
  calculate(stat = "diff in means", order = c("Winter", "Rest of Year"))

obs_stat_postcovid
```

## Step4:  Generate null distribution and visualise
```{r}
null_dist_postcovid <- ae_winter_postcovid %>% 
  specify(response = NumberOfAttendancesAggregate, explanatory = winter) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("Winter", "Rest of Year"))

```

```{r}
null_dist_postcovid %>% 
  visualise(bins = 25)+
  shade_p_value(obs_stat = obs_stat_postcovid$stat, 
                direction = "greater")
```

## Step 5: Get p-value and interpret


```{r}
null_dist_postcovid %>% 
  get_p_value(obs_stat = obs_stat_precovid$stat, 
                direction = "greater")
```

Fail to reject the null hypothesis, ie the number of admissions to A&E in winter is the same as, or less than, admissions in the rest of the year. 



```{r}
a_e_winter_covid %>% 
  group_by(winter, covid)
```

These figures indicate that pre and post covid there are always fewer mean admissions during winter, compared to the rest of the year. 

# Check on Test 2. 

Mean discharge to home rate is 1944.696. Is the discharge to home rate in the winter different to the mean for hte rest of the year? One -sample mean

## Step1: Generate hypothesis
H0: mean(winter discharge home) = mean(rest of year discharge home)
$$H_0:  \mu_{\textrm{winter discharge home}} =1944.696$$
H1: higher mean discharge home in rest of year compared to winter )
$$H_1:  \mu_{\textrm{winter discharged home}}\ne 1944.696$$
significance level = 0.05

## Step 2: calculate observed stat

```{r}
obs_stat <- a_e_clean_winter %>%
  filter(winter == "Winter") %>% 
  summarise(mean_discharge_home_winter = mean(DischargeDestinationResidence, na.rm = TRUE))

obs_stat
```

## Step 3: Generate null distribution


```{r}
winter_home_null_distr <- a_e_clean_winter %>% 
  specify(response = DischargeDestinationResidence) %>% 
  hypothesise(null = "point", mu = 1944.696) %>% 
  generate(reps = 10000, type = "bootstrap") %>% 
  calculate(stat = "mean")
```
## Step 4: Visualise


```{r}
winter_home_null_distr %>% 
  visualise(bins = 30) +
  shade_p_value(obs_stat = obs_stat$mean_discharge_home_winter, direction = "both")
```
## Step 5: Get p-value and interpret
```{r}
winter_home_null_distr %>% 
  get_p_value(obs_stat = obs_stat$mean_discharge_home_winter, direction = "both")
```


# Test 4: Are there differences in the mean proportions of people sent home in winter, compared to the rest of the year

## Step 1: generate hypothesis
Are the proportions of patients to A&E discharged to home different in winter and the rest of the year. (Two means - independent)

H0: mean(proportion home discharged rest of year) - mean(proportion home discharge winter) <= 0

$$H_0:  \mu_{\textrm{prop discharged rest of year}} -  \mu_{\textrm{prop discharge home winter }} <=0$$

H1: mean(proportion home discharge rest of year) - mean(proportion home discharge winter) > 0 

$$H_1:  \mu_{\textrm{prop discharged rest of year}} -  \mu_{\textrm{prop discharge home winter}}>0$$

H0: mean(proportion home discharged rest of year) - mean(proportion home discharge winter) <= 0
H1: mean(proportion home discharge rest of year) - mean(proportion home discharge winter) > 0 

## Data wrangling and explore (set up new prop_discharged_home variable)
```{r}
a_e_winter_covid <- a_e_winter_covid %>% 
  mutate(prop_discharged_home = DischargeDestinationResidence/ NumberOfAttendancesAggregate)
```


```{r}
a_e_winter_covid %>% 
  group_by(winter) %>% 
  summarise(mean_admissions = mean(NumberOfAttendancesAggregate),
            mean_discharge_home = mean(DischargeDestinationResidence, na.rm = TRUE),
            mean_transfer = mean(DischargeDestinationTransfer, na.rm = TRUE),
            mean_prop_discharge_home = mean(prop_discharged_home, na.rm = TRUE))
```
```{r}
a_e_winter_covid %>% 
  group_by(winter, covid) %>% 
  summarise(mean_admissions = mean(NumberOfAttendancesAggregate),
            mean_discharge_home = mean(DischargeDestinationResidence, na.rm = TRUE),
            mean_transfer = mean(DischargeDestinationTransfer, na.rm = TRUE),
            mean_prop_discharge_home = mean(prop_discharged_home, na.rm = TRUE))
```

```{r}
a_e_winter_covid %>% 
  drop_na(prop_discharged_home) %>% 
  ggplot(aes(x = prop_discharged_home, y = winter))+
  geom_boxplot()
```



## Step 2: Calculate observed stat

```{r}
obs_stat_prop <- a_e_winter_covid %>% 
  drop_na(prop_discharged_home) %>% 
  specify(response = prop_discharged_home, explanatory = winter) %>% 
  calculate(stat = "diff in means", order = c("Rest of Year", "Winter"))

obs_stat_prop
```
STep 3: calculate null distribution

```{r}
null_dist_prop <- a_e_winter_covid %>% 
  drop_na(prop_discharged_home) %>%
  specify(response = prop_discharged_home, explanatory = winter) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 5000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("Rest of Year", "Winter"))

```


```{r}
null_dist_prop %>% 
  visualise(bins = 25)+
  shade_p_value(obs_stat = obs_stat_prop$stat, 
                direction = "greater")
```

```{r}
null_dist_prop %>% 
  get_p_value(obs_stat = obs_stat_prop$stat, 
                direction = "greater")
```

The p-value is less than our 0.05 significance level, and therefore we can reject the H0 in facour of the Ha. This means that we the proportion of patients that are discharged home is higher in the rest of the year, compared to in winter time. 

























