---
title: "Covid-19 data cleaning script"
author: "Mandip Farmahan"
date: "2023-05-05"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

## Import libraries

```{r}
library(tidyverse)
library(here)
library(tsibble)
```

## Import data files

```{r, message=FALSE}
# Covid Hospital Admissions

hb_agesex <- read_csv(here("data/raw_data/covid_19/hospital_admissions_hb_agesex_20230504.csv"))
hb_specialty <- read_csv(here("data/raw_data/covid_19/hospital_admissions_hb_specialty_20230504.csv"))
hscp_agesex <- read_csv(here("data/raw_data/covid_19/hospital_admissions_hscp_agesex_20230504.csv"))
hscp_specialty <- read_csv(here("data/raw_data/covid_19/hospital_admissions_hscp_specialty_20230504.csv"))
```

```{r, message=FALSE}
# Helper files

hb_names <- read_csv(here("data/raw_data/covid_19/health_board_names.csv")) %>% 
  select(HB, HBName)
hscp_names<- read_csv(here("data/raw_data/covid_19/hscp_names.csv")) %>% 
  select(HSCP, HSCPName, HB, HBName)
qualifier_info <- read_csv(here("data/raw_data/covid_19/qualifier_info.csv"))
```

```{r}
tibble_cv_admissions <- tibble(
  "hb_agesex" = names(hb_agesex),
  "hb_specialty" = c(names(hb_specialty), 1:2),
  "hscp_agesex" = c(names(hscp_agesex), ""),
  "hscp_specialty" = c(names(hscp_specialty), 1:3)
)
```

<br>

----

# Data Analysis

## Admissions By Health Board, Age and Sex

```{r}
head(hb_agesex)

hb_agesex %>% 
  filter(WeekEnding == 20200105,
         HB == "S08000015") %>% 
  summarise(count = sum(NumberAdmissions, na.rm = TRUE))

hb_agesex %>% 
  count(HB, HBQF)

hb_agesex %>% 
  count(AgeGroup, AgeGroupQF)

hb_agesex %>% 
  count(Sex, SexQF)

hb_agesex %>% 
  count(AdmissionType, AdmissionTypeQF)
```

```{r}
hb_agesex_ts <- hb_agesex %>% 
  mutate(year = year(ymd(WeekEnding)),
         month = month(ymd(WeekEnding)),
         .after = WeekEnding
  ) %>% 
  mutate(AgeGroup = case_when(AgeGroup == "Under 5" ~ "0 - 05",
                              AgeGroup == "5 - 14" ~ "05 - 14",
                              TRUE ~ AgeGroup)) %>%
  mutate(pre_post_diff = NumberAdmissions - Average20182019) %>% 
  left_join(hb_names, by = "HB")
```

```{r, echo=FALSE}
# # count by AgeGroup
# hb_agesex_ts %>%
#   group_by(AgeGroup, year, month) %>% 
#   summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
#   ggplot(aes(x = month, y = num_admissions, colour = AgeGroup))+
#   geom_point()+
#   scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
#   facet_wrap(~ year, nrow = 1)
# 
# # count by AgeGroup -- removed unknown "All ages" group
# hb_agesex_ts %>%
#   filter(is.na(AgeGroupQF)) %>% 
#   group_by(AgeGroup, year, month) %>% 
#   summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
#   ggplot(aes(x = month, y = num_admissions, colour = AgeGroup))+
#   geom_line()+
#   scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
#   facet_wrap(~ year, nrow = 1)
# 
# # count by AgeGroup -- removed unknown "All ages" group (YEAR ONLY)
# hb_agesex_ts %>%
#   filter(is.na(AgeGroupQF)) %>% 
#   group_by(AgeGroup, year) %>% 
#   summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
#   ggplot(aes(x = year, y = num_admissions, colour = AgeGroup))+
#   geom_line()
# 
# # count by AgeGroup -- across each location (time unit = 1 week)
# hb_agesex_ts %>%
#   left_join(hb_names, by = "HB") %>% 
#   filter(!is.na(HBName),
#          is.na(AgeGroupQF)) %>% 
#   group_by(AgeGroup, HBName) %>% 
#   summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
#   ggplot(aes(x = AgeGroup, y = num_admissions, colour = HBName, group = HBName))+
#   geom_line()+
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r, echo=FALSE}
# # count difference from 2018/2019 by AgeGroup
# hb_agesex_ts %>%
#   filter(is.na(AgeGroupQF)) %>% 
#   group_by(AgeGroup, year, month) %>% 
#   summarise(num_admissions = mean(pre_post_diff, na.rm = TRUE)) %>% 
#   ggplot(aes(x = month, y = num_admissions, colour = AgeGroup))+
#   geom_point()+
#   scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
#   facet_wrap(~ year, nrow = 1)
# 
# # percentage difference from 2018/2019 by AgeGroup
# hb_agesex_ts %>%
#   filter(is.na(AgeGroupQF)) %>% 
#   group_by(AgeGroup, year, month) %>% 
#   summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
#   ggplot(aes(x = month, y = percentage_diff, colour = AgeGroup))+
#   geom_point()+
#   scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
#   facet_wrap(~ year, nrow = 1)
# 
# # percentage difference from 2018/2019 by AgeGroup 
# hb_agesex_ts %>%
#   filter(is.na(AgeGroupQF)) %>% 
#   group_by(AgeGroup, year, month) %>% 
#   summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
#   ggplot(aes(x = month, y = percentage_diff, colour = AgeGroup))+
#   geom_point()+
#   ylim(-50, 50)+ # smaller scale to focus on central area
#   scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
#   facet_wrap(~ year, nrow = 1)
# 
# # percentage difference from 2018/2019 by AgeGroup (YEAR ONLY)
# hb_agesex_ts %>%
#   filter(is.na(AgeGroupQF)) %>% 
#   group_by(AgeGroup, year) %>% 
#   summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
#   ggplot(aes(x = year, y = percentage_diff, colour = AgeGroup))+
#   geom_line()
# 
# # percentage difference from 2018/2019 by AgeGroup and location (time unit = 1 week)
# hb_agesex_ts %>%
#   left_join(hb_names, by = "HB") %>% 
#   filter(!is.na(HBName),
#          is.na(AgeGroupQF)) %>% 
#   group_by(AgeGroup, HBName) %>% 
#   summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
#   ggplot(aes(x = AgeGroup, y = percentage_diff, colour = HBName, group = HBName))+
#   geom_line()+
#   ylim(-50, 50)+ # smaller scale to focus on central area
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

### Age

Split by year (Weekly average number of admissions per year)

```{r}
# number of admissions-- MONTHLY CHECK FOR PATTERNS
hb_agesex_ts %>%
  filter(is.na(AgeGroupQF)) %>%
  group_by(AgeGroup, year, month) %>%
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = num_admissions, colour = AgeGroup))+
  geom_line()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(~ year, nrow = 1)
```


```{r}
# number of admissions
hb_agesex_ts %>%
  # filter(!is.na(HBName)) %>% # uncomment to include location
  filter(is.na(AgeGroupQF)) %>%
  group_by(AgeGroup, year) %>% 
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = num_admissions, colour = AgeGroup))+
  geom_line()

# percentage difference from 2018/2019
hb_agesex_ts %>%
  # filter(!is.na(HBName)) %>% # uncomment to include location
  filter(is.na(AgeGroupQF)) %>% 
  group_by(AgeGroup, year) %>% 
  summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = percentage_diff, colour = AgeGroup))+
  geom_line()
```

### Gender

Split by year (Weekly average number of admissions per year)

```{r}
# number of admissions -- MONTHLY CHECK FOR PATTERNS
hb_agesex_ts %>%
  filter(is.na(SexQF)) %>%
  group_by(Sex, year, month) %>%
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = num_admissions, colour = Sex))+
  geom_line()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(~ year, nrow = 1)
```


```{r}
# number of admissions
hb_agesex_ts %>%
  # filter(!is.na(HBName)) %>% # uncomment to include location
  filter(is.na(SexQF)) %>%
  group_by(Sex, year) %>% 
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = num_admissions, colour = Sex))+
  geom_line()

# percentage difference from 2018/2019
hb_agesex_ts %>%
  # filter(!is.na(HBName)) %>% # uncomment to include location
  filter(is.na(SexQF)) %>% 
  group_by(Sex, year) %>% 
  summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = percentage_diff, colour = Sex))+
  geom_line()
```

Summary: There doesn't seem to be a huge change pre and post covid outbreak, when compared across the ages. If anything, it looks like there is a decrease in admissions for most age groups, possibly as there were no beds available for non-covid cases. The exceptions are children (14 and under) and elderly (85 and over) in 2021 onwards.

The most notable data points are in April 2020 (huge dip), which I believe is when the news broke in the UK about covid as the cases first started being talked about in the media (March/April). I can't remember if we were put into lock down but WFH definitely started then started then. 

<br>

---

## Admissions By Health Board and Deprivation

file is empty

<br>

---

## Admissions By Health Board and Specialty

```{r}
head(hb_specialty)

hb_specialty %>% 
  filter(WeekEnding == 20200105,
         HB == "S08000015") %>% 
  summarise(count = sum(NumberAdmissions, na.rm = TRUE))

hb_specialty %>% 
  count(AdmissionType, AdmissionTypeQF)

hb_specialty %>% 
  count(Specialty, SpecialtyQF)

```

```{r}
hb_specialty_ts <- hb_specialty %>% 
  mutate(year = year(ymd(WeekEnding)),
         month = month(ymd(WeekEnding)),
         .after = WeekEnding
  ) %>% 
  mutate(pre_post_diff = NumberAdmissions - Average20182019) %>% 
  left_join(hb_names, by = "HB")
```


```{r}
# number of admissions-- MONTHLY CHECK FOR PATTERNS
hb_specialty_ts %>%
  filter(is.na(SpecialtyQF)) %>%
  group_by(Specialty, AdmissionType, year, month) %>%
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = num_admissions, colour = Specialty))+
  geom_line()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(AdmissionType ~ year)

# number of admissions-- MONTHLY CHECK FOR PATTERNS - remove generic specialty
hb_specialty_ts %>%
  filter(is.na(SpecialtyQF)) %>%
  filter(!is.na(Specialty),
         !str_detect(Specialty, "^Medical")) %>% 
  group_by(Specialty, AdmissionType, year, month) %>%
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = num_admissions, colour = Specialty))+
  geom_line()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(AdmissionType ~ year)
```


```{r}
# number of admissions
hb_specialty_ts %>%
  # filter(!is.na(HBName)) %>% # uncomment to include location
  filter(is.na(SpecialtyQF)) %>%
  filter(!is.na(Specialty), # remover NAs
         !str_detect(Specialty, "^Medical")) %>% 
  group_by(Specialty, year) %>% 
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = num_admissions, colour = Specialty))+
  geom_line()

# percentage difference from 2018/2019
hb_specialty_ts %>%
  # filter(!is.na(HBName)) %>% # uncomment to include location
  filter(is.na(SpecialtyQF)) %>% 
  filter(!is.na(Specialty),  # remover NAs
         !str_detect(Specialty, "^Medical")) %>% 
  group_by(Specialty, year) %>% 
  summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = percentage_diff, colour = Specialty))+
  geom_line()
```

<br>

---

## Admissions By HSCP, Age and Sex

```{r}
head(hscp_agesex)

hscp_agesex %>% 
  filter(WeekEnding == 20200105,
         HSCP == "S37000001") %>% 
  summarise(count = sum(NumberAdmissions, na.rm = TRUE))
```

```{r}
hscp_agesex_ts <- hscp_agesex %>% 
  mutate(year = year(ymd(WeekEnding)),
         month = month(ymd(WeekEnding)),
         .after = WeekEnding
  ) %>% 
  mutate(AgeGroup = case_when(AgeGroup == "Under 5" ~ "0 - 05",
                              AgeGroup == "5 - 14" ~ "05 - 14",
                              TRUE ~ AgeGroup)) %>%
  mutate(pre_post_diff = NumberAdmissions - Average20182019) %>% 
  left_join(hscp_names, by = "HSCP")
```

### Age

Split by year (Weekly average number of admissions per year)

```{r}
# number of admissions-- MONTHLY CHECK FOR PATTERNS
hscp_agesex_ts %>%
  filter(is.na(AgeGroupQF)) %>%
  group_by(AgeGroup, year, month) %>%
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = num_admissions, colour = AgeGroup))+
  geom_line()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(~ year, nrow = 1)
```


```{r}
# number of admissions
hscp_agesex_ts %>%
  # filter(!is.na(HSCPName)) %>% # uncomment to include location
  filter(is.na(AgeGroupQF)) %>%
  group_by(AgeGroup, year) %>% 
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = num_admissions, colour = AgeGroup))+
  geom_line()

# percentage difference from 2018/2019
hscp_agesex_ts %>%
  # filter(!is.na(HSCPName)) %>% # uncomment to include location
  filter(is.na(AgeGroupQF)) %>% 
  group_by(AgeGroup, year) %>% 
  summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = percentage_diff, colour = AgeGroup))+
  geom_line()
```

### Gender

Split by year (Weekly average number of admissions per year)

```{r}
# number of admissions -- MONTHLY CHECK FOR PATTERNS
hscp_agesex_ts %>%
  filter(is.na(SexQF)) %>%
  group_by(Sex, year, month) %>%
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = num_admissions, colour = Sex))+
  geom_line()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(~ year, nrow = 1)
```


```{r}
# number of admissions
hscp_agesex_ts %>%
  # filter(!is.na(HSCPName)) %>% # uncomment to include location
  filter(is.na(SexQF)) %>%
  group_by(Sex, year) %>% 
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = num_admissions, colour = Sex))+
  geom_line()

# percentage difference from 2018/2019
hscp_agesex_ts %>%
  # filter(!is.na(HSCPName)) %>% # uncomment to include location
  filter(is.na(SexQF)) %>% 
  group_by(Sex, year) %>% 
  summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = percentage_diff, colour = Sex))+
  geom_line()
```


<br>

---

## Admissions By HSCP and Deprivation

file is empty

<br>

---

## Admissions By HSCP and Specialty

```{r}
head(hscp_specialty)

hscp_specialty %>% 
  filter(WeekEnding == 20200105,
         HSCP == "S37000001") %>% 
  summarise(count = sum(NumberAdmissions, na.rm = TRUE))
```

```{r}
hscp_specialty_ts <- hscp_specialty %>% 
  mutate(year = year(ymd(WeekEnding)),
         month = month(ymd(WeekEnding)),
         .after = WeekEnding
  ) %>% 
  mutate(pre_post_diff = NumberAdmissions - Average20182019) %>% 
  left_join(hscp_names, by = "HSCP")
```


```{r}
# number of admissions-- MONTHLY CHECK FOR PATTERNS
hscp_specialty_ts %>%
  filter(is.na(SpecialtyQF)) %>%
  group_by(Specialty, AdmissionType, year, month) %>%
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = num_admissions, colour = Specialty))+
  geom_line()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(AdmissionType ~ year)

# number of admissions-- MONTHLY CHECK FOR PATTERNS - remove generic specialty
hscp_specialty_ts %>%
  filter(is.na(SpecialtyQF)) %>%
  filter(!is.na(Specialty),
         !str_detect(Specialty, "^Medical")) %>% 
  group_by(Specialty, AdmissionType, year, month) %>%
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = num_admissions, colour = Specialty))+
  geom_line()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(AdmissionType ~ year)
```


```{r}
# number of admissions
hscp_specialty_ts %>%
  # filter(!is.na(HSCPName)) %>% # uncomment to include location
  filter(is.na(SpecialtyQF)) %>%
  filter(!is.na(Specialty), # remover NAs
         !str_detect(Specialty, "^Medical")) %>% 
  group_by(Specialty, year) %>% 
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = num_admissions, colour = Specialty))+
  geom_line()

# percentage difference from 2018/2019
hscp_specialty_ts %>%
  # filter(!is.na(HSCPName)) %>% # uncomment to include location
  filter(is.na(SpecialtyQF)) %>% 
  filter(!is.na(Specialty),  # remover NAs
         !str_detect(Specialty, "^Medical")) %>% 
  group_by(Specialty, year) %>% 
  summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = percentage_diff, colour = Specialty))+
  geom_line()
```

<br>

---






























