---
title: "Covid-19 data cleaning script"
author: "Mandip Farmahan"
date: "2023-05-05"
output: html_document
---

## Import libraries

```{r}
library(tidyverse)
library(here)
library(tsibble)
```

## Import data files

```{r, message=FALSE}
# Covid Hospital Admissions

hb_agesex <- read_csv(here("data/raw_data/covid_19/hospital_admissions_hb_agesex_20230504.csv"))
hb_specialty <- read_csv(here("data/raw_data/covid_19/hospital_admissions_hb_specialty_20230504.csv"))
hscp_agesex <- read_csv(here("data/raw_data/covid_19/hospital_admissions_hscp_agesex_20230504.csv"))
hscp_specialty <- read_csv(here("data/raw_data/covid_19/hospital_admissions_hscp_specialty_20230504.csv"))
```

```{r, message=FALSE}
# Helper files

hb_names <- read_csv(here("data/raw_data/covid_19/health_board_names.csv"))
qualifier_info <- read_csv(here("data/raw_data/covid_19/qualifier_info.csv"))
```

```{r}
tibble_cv_admissions <- tibble(
  "hb_agesex" = names(hb_agesex),
  "hb_specialty" = c(names(hb_specialty), 1:2),
  "hscp_agesex" = c(names(hscp_agesex), ""),
  "hscp_specialty" = c(names(hscp_specialty), 1:3)
)
```

<br>

----

# Data Analysis

## Admissions By Health Board, Age and Sex

```{r}
head(hb_agesex)

hb_agesex %>% 
  count(HBQF)

hb_agesex %>% 
  count(AgeGroup, AgeGroupQF)

hb_agesex %>% 
  count(Sex, SexQF)

hb_agesex %>% 
  count(AdmissionType, AdmissionTypeQF)
```

```{r}
hb_agesex_ts <- hb_agesex %>% 
  mutate(year = year(ymd(WeekEnding)),
         month = month(ymd(WeekEnding)),
         .after = WeekEnding
         ) %>% 
  mutate(pre_post_diff = NumberAdmissions - Average20182019)

```

```{r}
# count by AgeGroup
hb_agesex_ts %>%
  group_by(AgeGroup, year, month) %>% 
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
  ggplot(aes(x = month, y = num_admissions, colour = AgeGroup))+
  geom_point()+
    scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(~ year, nrow = 1)

# count by AgeGroup -- removed unknown "All ages" group
hb_agesex_ts %>%
  filter(is.na(AgeGroupQF)) %>% 
  group_by(AgeGroup, year, month) %>% 
  summarise(num_admissions = mean(NumberAdmissions, na.rm = TRUE)) %>% 
  ggplot(aes(x = month, y = num_admissions, colour = AgeGroup))+
  geom_line()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(~ year, nrow = 1)
```


```{r}
# count difference from 2018/2019 by AgeGroup
hb_agesex_ts %>%
  filter(is.na(AgeGroupQF)) %>% 
  group_by(AgeGroup, year, month) %>% 
  summarise(num_admissions = mean(pre_post_diff, na.rm = TRUE)) %>% 
  ggplot(aes(x = month, y = num_admissions, colour = AgeGroup))+
  geom_point()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(~ year, nrow = 1)

# percentage difference from 2018/2019 by AgeGroup
hb_agesex_ts %>%
  filter(is.na(AgeGroupQF)) %>% 
  group_by(AgeGroup, year, month) %>% 
  summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
  ggplot(aes(x = month, y = percentage_diff, colour = AgeGroup))+
  geom_point()+
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(~ year, nrow = 1)

# percentage difference from 2018/2019 by AgeGroup 
hb_agesex_ts %>%
  filter(is.na(AgeGroupQF)) %>% 
  group_by(AgeGroup, year, month) %>% 
  summarise(percentage_diff = mean(PercentVariation, na.rm = TRUE)) %>% 
  ggplot(aes(x = month, y = percentage_diff, colour = AgeGroup))+
  geom_point()+
  ylim(-50, 50)+ # smaller scale to focus on central area
  scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11))+
  facet_wrap(~ year, nrow = 1)
```

Summary: There doesn't seem to be a huge change pre and post covid outbreak, when compared across the ages. If anything, it looks like there is a decrease in admissions for most age groups, possibly as there were no beds available for non-covid cases. The exceptions are children (14 and under) and elderly (85 and over) in 2021 onwards.

The most notable data points are in April 2020 (huge dip), which I believe is when the news broke in the UK about covid as the cases first started being talked about in the media (March/April). I can't remember if we were put into lock down but WFH definitely started then started then. 

<br>

---

## Admissions By Health Board and Deprivation

file is empty

<br>

---

## Admissions By Health Board and Specialty

```{r}
head(hb_specialty)

```

<br>

---

## Admissions By HSCP, Age and Sex

```{r}
head(hscp_agesex)
```

<br>

---

## Admissions By HSCP and Deprivation

file is empty

<br>

---

## Admissions By HSCP and Specialty

```{r}
head(hscp_specialty)
```

<br>

---






























